---
title: "Aggregations with agg_by"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{agg_by}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Table aggregations are a quintessential feature of Deephaven. You can apply as
many aggregations as needed to static tables _or_ streaming tables, and if the
parent tables are streaming, the resulting aggregated tables will update
alongside their parent tables. It is also very easy to perform _grouped_
aggregations, which allow you to aggregate tables on a per-group basis.

## Applying aggregations to a table

There are two methods for performing aggregations on a table, `agg_by()` and
`agg_all_by()`. `agg_by()` allows you to perform many aggregations on specified
columns, while `agg_all_by()` allows you to perform a single aggregation to
every non-grouping column in the table. Both methods have an optional `by`
parameter that is used to specify grouping columns. Here are some details on
each method:

- `TableHandle$agg_by(aggs, by)`: Creates a new table containing grouping
  columns and grouped data.
  The resulting grouped data is defined by the aggregation(s) specified.
- `TableHandle$agg_all_by(agg, by)`: Creates a new table containing grouping
  columns and grouped data. The resulting grouped data is defined by the
  aggregation specified. This method applies the aggregation to all non-grouping
  columns of the table, so it can only accept one aggregation at a time.

The `agg_by()` and `agg_all_by()` methods themselves do not know anything about
the columns on which you want to perform aggregations. Rather, the desired
columns are passed to individual `agg` functions, enabling you to apply various
kinds of aggregations to different columns or groups of columns as needed.

## `agg` functions

`agg` functions are used to perform aggregation calculations on grouped data by
passing them to `agg_by()` or `agg_all_by()`. These functions are _generators_,
meaning they return _functions_ that the Deephaven engine knows how to
interpret. We call the functions that they return `AggOp`s. See `?AggOp` for
more information. These AggOps are not R-level functions, but
Deephaven-specific data types that perform all of the intensive calculations.
Here is a list of all `agg` functions available in Deephaven:

- `agg_first()`
- `agg_last()`
- `agg_min()`
- `agg_max()`
- `agg_sum()`
- `agg_abs_sum()`
- `agg_avg()`
- `agg_w_avg()`
- `agg_median()`
- `agg_var()`
- `agg_std()`
- `agg_percentile()`
- `agg_count()`

For more details on each aggregation function, see the reference documentation
by running `?agg_first`, `?agg_last`, etc.

## An Example
```{r, eval=FALSE}
library(rdeephaven)

# connecting to Deephaven server
client <- Client$new("localhost:10000", auth_type = "psk", auth_token = "my_secret_token")

# create data frame, push to server, retrieve TableHandle
df <- data.frame(
  X = c("A", "B", "A", "C", "B", "A", "B", "B", "C"),
  Y = c("M", "N", "O", "N", "P", "M", "O", "P", "M"),
  Number1 = c(100, -44, 49, 11, -66, 50, 29, 18, -70),
  Number2 = c(-55, 76, 20, 130, 230, -50, 73, 137, 214)
)
th <- client$import_table(df)

# get first and last elements of each column
th1 <- th$
  agg_by(
  agg_first(c("XFirst = X", "YFirst = Y", "Number1First = Number1", "Number2First = Number2")),
  agg_last(c("XLast = X", "YLast = Y", "Number1Last = Number1", "Number2Last = Number2"))
)

# compute mean and standard deviation of Number1 and Number2, grouped by X
th2 <- th$
  agg_by(
  c(
    agg_avg(c("Number1Avg = Number1", "Number2Avg = Number2")),
    agg_std(c("Number1Std = Number1", "Number2Std = Number2"))
  ),
  by = "X"
)

# compute maximum of all non-grouping columns, grouped by X and Y
th3 <- th$
  agg_all_by(agg_max(), by = c("X", "Y"))

# compute minimum and maximum of Number1 and Number2 respectively grouped by Y
th4 <- th$
  agg_by(
  c(
    agg_min("Number1Min = Number1"),
    agg_max("Number2Max = Number2")
  ),
  by = "Y"
)

client$close()
```
