---
title: "Overview of the Deephaven Core R Client"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{rdeephaven}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The Deephaven Core R Client provides an R interface to Deephaven's powerful
real-time data engine, [_Deephaven Core_](https://deephaven.io/community/).
To use this package, you must have a Deephaven server running and be able to 
connect to it. For more information on how to set up a Deephaven server, see the 
documentation [here](https://deephaven.io/core/docs/tutorials/quickstart/).

## Building blocks of the Deephaven R Client

There are two primary R classes that make up the Deephaven R Client, the `Client` 
class and the `TableHandle` class. See `?Client` and `?TableHandle` for more
information. The Client class is used to establish a connection to the
Deephaven server with its constructor `Client$new()`, and to send server 
requests, such as running a script via `run_script()`, or pushing local data to 
the server via `import_table()`. Basic usage of the Client class may look 
something like this:

```{r, eval=FALSE}
library(rdeephaven)
client <- Client$new("localhost:10000", auth_type = "psk", auth_token = "my_secret_token")
```

Many of these server requests end up creating or modifying tables that live on 
the server. To keep track of these tables, the R client retrieves references to 
them, and wraps these references in TableHandle objects. These TableHandles 
have a host of methods that mirror server-side table operations, such as
`head()`, `tail()`, `update_by()`, and so on. So, you can typically use
TableHandles _as if_ they are tables themselves, and all of the corresponding
methods that you call on them will be executed on the server. Here is a simple
example of pushing data to the server, retrieving a TableHandle to the resulting
table, and applying some basic table operations to the table:

```{r, eval=FALSE}
df1 <- data.frame(x = 1:10, y = 11:20)
th1 <- client$import_table(df1)

th2 <- th1$
  update("z = x + y")$
  where("z % 4 == 0")
```

TableHandles also support common functional methods for converting server-side 
Deephaven tables to R objects stored in local memory such as `as.data.frame()`, 
`as_tibble()`, and `as_arrow_table()`. Here's an example of converting the table 
created above to an R data frame and verifying that other functional methods
work as expected:

```{r, eval=FALSE}
df2 <- as.data.frame(th2)

print(nrow(th2) == nrow(df2))
print(ncol(th2) == ncol(df2))
print(dim(th2) == dim(df2))
print(all(as.data.frame(head(th2, 2)) == head(df2, 2)))
print(all(as.data.frame(tail(th2, 2)) == tail(df2, 2)))
```

For more information on these classes and all of their methods, see the
reference documentation by running `?Client` or `?TableHandle`.

## Real-time data analysis

Since TableHandles are references to tables living on the Deephaven server, they 
may refer to streaming tables, or tables that are receiving new data
periodically(typically once per second). Here's a simple example of creating a
table that adds a new row every second:

```{r, eval=FALSE}
th3 <- client$time_table("PT1s")$
  update(c("X = ii", "Y = sin(X)"))
```

R objects like data frames or Dplyr tibbles do not have this streaming property 
- they are always static objects stored in memory. However, a TableHandle 
referring to a streaming table may be converted to a data frame or tibble at any 
time, and the resulting object will be a snapshot of the table at the time of 
conversion. This means that you can use the Deephaven R Client to perform
real-time data analysis on streaming data! Here, we make a simple plot of the
ticking table, and call it three times to demonstrate the dynamic nature of the
table:

```{r, eval=FALSE}
plot(as.data.frame(th3)$X, as.data.frame(th3)$Y, type = "l")
Sys.sleep(5)
plot(as.data.frame(th3)$X, as.data.frame(th3)$Y, type = "l")
Sys.sleep(5)
plot(as.data.frame(th3)$X, as.data.frame(th3)$Y, type = "l")
```

There are performance and memory considerations when pulling data from the
server, so it is best to use the provided TableHandle methods to perform as much
of your analysis as possible on the server, and to only pull the data when
something _must_ be done in R, like plotting or writing to a local file.

## Powerful table operations

Much of the power of Deephaven's suite of table operations is achieved through 
the use of the `update_by()` and `agg_by()` methods. These table methods are
important enough to warrant their own documentation pages, accessible by running
`vignette("update_by")` or `vignette("agg_by")`. These methods
come with their own suites of functions, prefixed with `agg_` and `uby_`
respectively, that are listed in their documentation pages. Running
`ls("package:rdeephaven")` will reveal that most of the functions included in
this package are for these methods, so it is important to get acquainted with
them.

## Getting help
While we've done our best to provide good documentation for this package, you
may find you need more help than what this documentation has to offer. Please
visit the official Deephaven Community Core
[documentation](https://deephaven.io/core/docs/tutorials/quickstart/) to learn
more about Deephaven and to find comprehensive examples. Additionally, feel free
to reach out to us on the Deephaven
[Community Slack channel](https://deephaven.io/slack) with any questions. We
hope you find real-time data analysis in R to be as easy as possible.
