---
title: "Moving Operations with update_by"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{update_by}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Deephaven's `update_by()` table method and suite of `uby` functions enable
cumulative and moving calculations on static _and_ streaming tables. Complex
operations like cumulative minima and maxima, exponential moving averages, and
rolling standard deviations are all possible and effortless to execute. As
always in Deephaven, the results of these calculations will continue to update
as their parent tables are updated. Additionally, it's easy to group data by one
or more columns, enabling complex group-wise calculations with a single line of
code.

## Applying UpdateBy operations to a table

The table method `update_by()` is the entry point for UpdateBy operations. It
takes two arguments: the first is an `UpdateByOp` or a list of
`UpdateByOp`s denoting the calculations to perform on specific columns of the
table. Then, it takes a column name or a list of column names that define the
groups on which to perform the calculations. If you don't want grouped
calculations, omit this argument.

To learn more about UpdateByOps, see the reference documentation with
`?UpdateByOp`.

The `update_by()` method itself does not know anything about the columns on
which you want to perform calculations. Rather, the desired columns are passed
to individual `uby` functions, enabling a massive amount of flexibility.

## `uby` functions

`uby` functions are the workers that actually execute the complex UpdateBy
calculations. These functions are _generators_, meaning they return _functions_
that the Deephaven engine knows how to interpret. We call the functions they
return `UpdateByOp`s. See `?UpdateByOp` for more information. These UpdateByOps
are not R-level functions, but Deephaven-specific data types that perform all of
the intensive calculations. Here is a list of all `uby` functions available in
Deephaven:

- `uby_cum_min()`
- `uby_cum_max()`
- `uby_cum_sum()`
- `uby_cum_prod()`
- `uby_forward_fill()`
- `uby_delta()`
- `uby_emmin_tick()`
- `uby_emmin_time()`
- `uby_emmax_tick()`
- `uby_emmax_time()`
- `uby_ems_tick()`
- `uby_ems_time()`
- `uby_ema_tick()`
- `uby_ema_time()`
- `uby_emstd_tick()`
- `uby_emstd_time()`
- `uby_rolling_count_tick()`
- `uby_rolling_count_time()`
- `uby_rolling_group_tick()`
- `uby_rolling_group_time()`
- `uby_rolling_min_tick()`
- `uby_rolling_min_time()`
- `uby_rolling_max_tick()`
- `uby_rolling_max_time()`
- `uby_rolling_sum_tick()`
- `uby_rolling_sum_time()`
- `uby_rolling_prod_tick()`
- `uby_rolling_prod_time()`
- `uby_rolling_avg_tick()`
- `uby_rolling_avg_time()`
- `uby_rolling_wavg_tick()`
- `uby_rolling_wavg_time()`
- `uby_rolling_std_tick()`
- `uby_rolling_std_time()`

For more details on each aggregation function, see the reference documentation
by running `?uby_cum_min`, `?uby_delta`, etc.

## An Example
```{r, eval=FALSE}
library(rdeephaven)

# connecting to Deephaven server
client <- Client$new("localhost:10000", auth_type = "psk", auth_token = "my_secret_token")

# create data frame, push to server, retrieve TableHandle
df <- data.frame(
  timeCol = seq.POSIXt(as.POSIXct(Sys.Date()), as.POSIXct(Sys.Date() + 0.01), by = "1 sec")[1:500],
  boolCol = sample(c(TRUE, FALSE), 500, TRUE),
  col1 = sample(10000, size = 500, replace = TRUE),
  col2 = sample(10000, size = 500, replace = TRUE),
  col3 = 1:500
)
th <- client$import_table(df)

# compute 10-row exponential weighted moving average of col1 and col2, grouped by boolCol
th1 <- th$
  update_by(uby_ema_tick(decay_ticks = 10, cols = c("col1Ema = col1", "col2Ema = col2")), by = "boolCol")

# compute rolling 10-second weighted average and standard deviation of col1 and col2, weighted by col3
th2 <- th$
  update_by(
  c(
    uby_rolling_wavg_time(ts_col = "timeCol", wcol = "col3", cols = c("col1WAvg = col1", "col2WAvg = col2"), rev_time = "PT10s"),
    uby_rolling_std_time(ts_col = "timeCol", cols = c("col1Std = col1", "col2Std = col2"), rev_time = "PT10s")
  )
)

# compute cumulative minimum and maximum of col1 and col2 respectively, and the rolling 20-row sum of col3, grouped by boolCol
th3 <- th$
  update_by(
  c(
    uby_cum_min(cols = "col1"),
    uby_cum_max(cols = "col2"),
    uby_rolling_sum_tick(cols = "col3", rev_ticks = 20)
  ),
  by = "boolCol"
)

client$close()
}
```